Quantifying and controlling biases in dark matter halo concentration estimates

### 2016-09-07 paper/ms.tex
Correcciones de revision listas

\begin{abstract}
We use bootstrapping to estimate the bias of concentration
estimates on N-body dark matter halos as a function of particle
number. 
We find that algorithms based on the maximum radial velocity
and radial particle binning tend to overestimate the concentration by 
$15\%-20\%$ for halos sampled with $200$ particles and by $7\%$-$10\%$
for halos sampled with $500$ particles. 
To control this bias at low particle numbers we propose a new
algorithm that estimates halo concentrations based on the integrated
mass profile.
The method uses the full particle information without any binning,
making it reliable in cases when low numerical resolution becomes a
limitation for other methods.
This method reduces the bias to $< 3\%$ for halos sampled with
$200$-$500$ particles.
The velocity and density methods have to use halos with least $\sim
4000$ particles in order to keep the biases down to the same low
level. 
We also show that the mass-concentration relationship could be
shallower than expected once the biases  of the different
concentration measurements are taken into account. 
These results show that bootstrapping and the concentration estimates
based on the integrated mass profile are valuable tools to probe the
internal structure of dark matter halos in numerical simulations. 
\end{abstract}

@@

The internal DM distribution in a halo is usually parameterized
through the density profile.  
In a first approximation this profile is spherically symmetric; the
density only depends on the radial coordinate.  
One of the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.   
This profile can be considered as universal \citep{Navarro2010}, assuming that 
one is not interested in the very central region where galaxy
formation takes place, and where the effects of baryon physics on the
dark matter distribution are still unknown. 
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides a conceptual framework
to study simulated dark matter halos as a function of redshift and
cosmological parameters.  
Numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014,Ludlow2016,Klypin2016}
summarized their results through the mass-concentration
relationship; that is, the distribution of concentration values at a
fixed halo mass and redshift.  
The success of such numerical experiments rests on a reliable
algorithm to estimate the concentration.  
Such an algorithm should provide unbiased results and must be robust
when applied at varying numerical resolution.

@@

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density, $H$
is the Hubble constant, $G$ and the universal gravitational constant,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius.  This radius marks the point where the
logarithmic slope of the density profile is equal to -2, the
transition between the power law scaling $\rho\propto r^{-1}$ for
$r<r_s$ and $\rho\propto r^{-3}$ for $r>r_s$.

@@

To date, there are two standard methods to estimate concentrations in
dark matter halos extracted from N-body simulations.  The first method
takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  Then, it estimates the
density in each logarithmic bin.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of dark matter halos. 
A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).

@@

We use two different simulations to test our methods.
The first is the Bolshoi run, a cosmological simulation that follows
the non-linear evolution of a dark matter density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.   The
cosmological parameters use a Hubble parameter $h=0.73$, a matter
density $\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The data is publicly available at \url{http://www.cosmosim.org/}.  
Details about the structure of the database and the simulation can be
found in \citep{2011ApJ...740..102K,2013AN....334..691R}.

@@

The right panel in Figure \ref{fig:downsampling} shows different uncertainty
results.
The lines show the difference between the 14 and 86 percentiles in the
$f_{\rm off}$ distribution at fixed mass. 
Each line corresponds to the three different methods to estimate the
concentration and the two simulations.
This shows that the bootstrapping technique can help us to assign a
$1\sigma$ uncertainty to the concentration values at a fixed $N_s$as 
%
\begin{equation}
\sigma_{c} = \frac{0.40}{\sqrt{N_s/200}}.
\end{equation}
%

@@

 We acknowledge financial support from Uniandes
 and Universidad de Antioquia. 
 We thank Tom\'as Verdugo, Stefan Gottloeber and Nelson  Padilla for
their feedback.
We thank the anonymous referees for comments that improved
the presentation of these results.

>>>>

\begin{abstract}
We use bootstrapping to estimate the bias of concentration estimates
on N-body dark matter halos as a function of particle number.  We find
that algorithms based on the maximum radial velocity and radial
particle binning tend to overestimate the concentration by $15\%-20\%$
for halos sampled with $200$ particles and by $7\%$-$10\%$ for halos
sampled with $500$ particles.  To control this bias at low particle
numbers we propose a new algorithm that estimates halo concentrations
based on the integrated mass profile.  The method uses the full
particle information without any binning, making it reliable in cases
when low numerical resolution becomes a limitation for other methods.
This method reduces the bias to $< 3\%$ for halos sampled with
$200$-$500$ particles.  The velocity and density methods have to use
halos with at least $\sim 4000$ particles in order to keep the biases
down to the same low level.  We also show that the mass-concentration
relationship could be shallower than expected once the biases of the
different concentration measurements are taken into account.  These
results show that bootstrapping and the concentration estimates based
on the integrated mass profile are valuable tools to probe the
internal structure of dark matter halos in numerical simulations.
\end{abstract}

@@

The internal DM distribution in a halo is usually parameterized
through the density profile.  
In a first approximation this profile is spherically symmetric; the
density only depends on the radial coordinate.  
One of the most popular radial parameterizations is the
Navarro-Frenk-White (NFW) profile \citep{NFW}.   
This profile can be considered as universal \citep{Navarro2010}, assuming that 
one is not interested in the very central region where galaxy
formation takes place, and where the effects of baryon physics on the
DM distribution are still unknown. 
This profile is a double power law in radius, where the transition
break happens at the so-called scale radius, $r_s$.  
The ratio between the scale radius and the halo virial radius $R_v$ is
known as the concentration $c=R_v/r_s$.

The concentration of the NFW profile provides a conceptual framework
to study simulated DM halos as a function of redshift and
cosmological parameters.  
Numerical studies
\citep{Neto2007,Maccio2008,Duffy2008,Munoz2011,Prada2012,Ludlow2014,Ludlow2016,Klypin2016}
summarized their results through the mass-concentration
relationship; that is, the distribution of concentration values at a
fixed halo mass and redshift.  
The success of such numerical experiments rests on a reliable
algorithm to estimate the concentration.  
Such an algorithm should provide unbiased results and must be robust
when applied at varying numerical resolution.

@@

\begin{equation}
\rho(r) = \frac{\rho_c\delta_c}{r/r_s(1+r/r_s)^2},
\label{eq:definition}
\end{equation}
%
where $\rho_c\equiv 3H^2/8\pi G$ is the Universe critical density, $H$
is the Hubble constant, $G$ the universal gravitational constant,
$\delta_c$ is the halo dimensionless characteristic density and $r_s$
is the scale radius.  This radius marks the point where the
logarithmic slope of the density profile is equal to -2, the
transition between the power law scaling $\rho\propto r^{-1}$ for
$r<r_s$ and $\rho\propto r^{-3}$ for $r>r_s$.

@@

To date, there are two standard methods to estimate concentrations in
DM halos extracted from N-body simulations.  The first method
takes all the particles in the halo and bins them in the logarithm of
the radial coordinate from the halo center.  Then, it estimates the
density in each logarithmic bin.  At this point is possible to make a
direct fit to the density as a function of the radial coordinate.
This method has been broadly used for more than two decades to study
the mass-concentration-redshift relation of DM halos. 
A second method uses the circular velocity profile.  It finds the
value of $x$ for which the normalized circular velocity $v(<x)$ shows
a maximum.  Using this value it solves numerically for the
corresponding value of the concentration using Eq. (\ref{eq:max_v}).

@@

We use two different simulations to test our methods.
The first is the Bolshoi run, a cosmological simulation that follows
the non-linear evolution of a DM density field sampled with
$2048^3$ particles over a cubic box of $250\ \hMpc$ on a side.   The
cosmological parameters use a Hubble parameter $h=0.73$, a matter
density $\Omega_m=0.3071$ and a normalization of the power spectrum
$\sigma_8=0.82$. 
The data is publicly available at \url{http://www.cosmosim.org/}.  
Details about the structure of the database and the simulation can be
found in \citep{2011ApJ...740..102K,2013AN....334..691R}.

@@

The right panel in Figure \ref{fig:downsampling} shows different
uncertainty results.  The lines show the difference between the 14 and
86 percentiles in the $f_{\rm off}$ distribution at fixed mass.  Each
line corresponds to the three different methods to estimate the
concentration applied to both simulations.  This shows that the
bootstrapping technique can help us to assign a $1\sigma$ uncertainty
to the concentration values at a fixed $N_s$as
%
\begin{equation}
\sigma_{c} = \frac{0.40}{\sqrt{N_s/200}}.
\end{equation}
%

@@

 We acknowledge financial support from Uniandes and Estrategia de
 Sostenibilidad 2014-2015 Universidad de Antioquia.  We thank Tom\'as
 Verdugo, Stefan Gottloeber and Nelson Padilla for their feedback.  We
 thank the anonymous referees for comments that improved the
 presentation of these results.
